---
title: "Project 2"
author: "Josh Stim"
date: "2023-09-24"
output: html_document
---

# Part 0: Setting up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install tidyverse and tidytuesdayR packages
```{r, echo = FALSE}
install.packages("tidyverse", "tidytuesdayR")
```

Load tidyverse and tidytuesdayR packages
```{r}
library(tidyverse)
library(tidytuesdayR)
```

# 1. Fun with functions

## Part a.

Write a function that computes the exponential of a number using the truncated series expansion. The function should take two arguments:

`x`: the number to be exponentiated

`k`: the number of terms to be used in the series expansion beyond the constant 1. The value of `k` is always $\geq 1$

Below is a recursive function that uses k terms to approximate exp(x)
```{r}
Exp <- function(x, k) {
  if (k == 0) {
    return(1)
  } else {
    return(Exp(x,k-1) + (x**k / factorial(k)))
  }
}
```

Below is output of my custom Exponential function using x = 2 and k = 15. The output is compared to exp(2).
```{r}
data.frame(Exp.output = Exp(2,20), exp.output = exp(2)) %>%
  mutate(difference = Exp.output - exp.output)
```


